### 2024-02-20 paper_sisc_first_revision/localpsf.tex
"Here," instead of "Here "

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. By ``non-negative integral kernel,'' we mean that entries of $\Aop$'s integral kernel 
%(defined in Equation \ref{eq:kernel_representation})
are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$. Such operators appear, for instance, as Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{BorziSchulz11,Delosreyes15,HinzePinnauUlbrichUlbrich08}, 
%AntilEtAl18
Schur complements in Schur complement methods for solving partial differential equations and Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps)~\cite{ChanMathew94,SmithBjorstadGropp96,ToselliWidlund04}, covariance operators in spatial statistics~\cite{ChenStein21,GentonKeyesTurkiyyah18,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}. Here, ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require solving a subproblem that involves PDEs, or performing other costly computations. 
%Matrix-free does not mean that matrices are never used; our approximation is ultimately converted into a hierarchical matrix (H-matrix), a compressed matrix format that permits fast and scalable linear algebra operations.

\begin{figure}
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[scale=0.3]{impulse_batch1.png}
		\caption{One impulse response batch}
		\label{fig:batches_intro}
	\end{subfigure}
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[scale=0.7]{mean_displacement_invariance.pdf}
		\caption{Local mean displacement invariance}
		\label{fig:mean_displacement_invariance}
	\end{subfigure}
%	\begin{subfigure}{0.32\textwidth}
%		\centering
%		\includegraphics[scale=0.34]{impulse_batch3.png}
%	\end{subfigure}
	\caption{(\ref{fig:batches_intro}) One batch, $\eta_b$, of normalized impulse responses, $\impulseresponse_{x}$, that arise from applying $\Aop$ to a weighted sum of scattered point sources (see Section \ref{sec:get_impulse_response}). Here $\Aop$ is the ice sheet inverse problem data misfit Gauss-Newton Hessian described in Section~\ref{sec:numerical_results}. 
	Black stars are point source locations. Shading shows the magnitude of the normalized impulse responses (darker means larger function values). Dashed gray ellipses are estimated impulse response support ellipsoids based on the moment method in Section \ref{sec:intromoments}. The large circle is $\partial \Omega$. (\ref{fig:mean_displacement_invariance}) Illustration of impulse responses, $\impulseresponse_x$ and $\impulseresponse_{x'}$, corresponding to points $x$ and $x'$. The operator $\Aop$ is locally mean displacement invariant (Section \ref{sec:local_mean_displacement_invariance}) if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.
	}
	\label{fig:batches_and_mdi}
\end{figure}

@@

Let $\febasis_1, \febasis_2, \dots, \febasis_\fedim$ be a set of continuous finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, let $V_h := \Span\left(\febasis_1, \febasis_2, \dots, \febasis_\fedim\right)$ be the corresponding finite element space under the $L^2$ inner product, and let $\fepoint_i \in \mathbb{R}^\gdim$, $i=1,\dots, \fedim$ be the Lagrange nodes associated with the functions $\febasis_i$. We assume that the finite element basis satisfies the Kronecker property, i.e., $\febasis_i(\fepoint_i)=1$ and $\febasis_i(\fepoint_j)=0$ if $i\neq j$. 
For $u_h \in V_h$ we write $\mathbf{u} \in \mathbb{R}^\nsamplepts_\massmatrix$ to denote the coefficient vector for $u_h$ with respect to the finite element basis, i.e.,
$
u_h(x) = \sum_{i=1}^\fedim \mathbf{u}_i \febasis_i(x).
$
Linear functionals $\rho_h \in V_h'$ have coefficient dual vectors $\boldsymbol{\rho}\in \mathbb{R}^\nsamplepts_{\massmatrix^{-1}}$, with entries $\boldsymbol{\rho}_i = \rho_h(\febasis_i)$ for $i=1,\dots,\nsamplepts$.
Here $\massmatrix \in \mathbb{R}^{\fedim \times \fedim}$ denotes the sparse finite element mass matrix which has entries $\massmatrix_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) \mathrm{d}x$ for $i,j=1,\dots,\fedim$.
The space $\mathbb{R}^\fedim_\massmatrix$ is $\mathbb{R}^\fedim$ with the inner product $(\mathbf{u},\mathbf{w})_\massmatrix := \mathbf{u}^T \massmatrix \mathbf{w}$, and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ is the analogous space with $\massmatrix^{-1}$ replacing $\massmatrix$. Direct calculation shows that $\mathbb{R}^\fedim_\massmatrix$ and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ are isomorphic to $V_h$ and $V_h'$ as Hilbert spaces, respectively.

@@

Let $C$, $L^i$, and $Q^{ij}$ be the following constant, linear, and quadratic functions:
\begin{equation*}
C(x) := 1, \qquad
L^i(x) := x^i, \qquad
Q^{ij}(x) := x^i x^j
\end{equation*}
for $i,j=1,\dots,\gdim$. Using the definition of $\spatialvol$ in~\eqref{eq:define_vol} and using~\eqref{eq:operator_simultaneous_ips}, we have
\begin{equation*}
	\spatialvol(x) = \int_\Omega \phi_x(y) C(y) ~ dy = \left(\phi_x, C\right)_{L^2(\Omega)} = \left(\mathcal{A}^T C\right)^*(x).
\end{equation*}
Hence, we compute $\spatialvol(x)$ for all $x$ simultaneously by applying $\Aop^T$ to $C$. Analogous manipulations show that $\spatialmean(x)$ and $\spatialcov(x)$ may be computed for all points $x$ simultaneously by applying $\Aop^T$ to the functions $L^i$ and $Q^{ij}$, respectively. We have
\begin{subequations}
	\label{eq:vol_mean_var}
	\begin{align}
	\spatialvol =& \left(\Aop^T C\right)^* \\
	\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
	\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
	\end{align}
\end{subequations}
for $i,j=1,\dots, \gdim$. Here $u/w$ denotes pointwise division, $\left(u/w\right)(x) = u(x)/w(x)$, and $u\cdot w$ denotes pointwise multiplication, $(u \cdot w)(x) = u(x)w(x)$.

@@

We start with a finite set of candidate points $\candidatepts$ and build $\pointbatch_k$ incrementally with points selected from $\candidatepts$. For simplicity of explanation, here $\pointbatch_k$ and $\candidatepts$ are mutable sets that we add points to and remove points from. First we initialize $\pointbatch_k$ as an empty set. Then we select the candidate point $\candidatepoint_i \in \candidatepts$ that is the farthest away from all points in previous sample point batches $S_1 \cup S_2 \cup \dots \cup S_{k-1}$. Candidate points for the first batch $S_1$ are chosen randomly from $\candidatepts$.
Once $\candidatepoint_i$ is selected, we remove $\candidatepoint_i$ from $\candidatepts$. Then we perform the following checks:
\begin{enumerate}
	\item We check whether $\candidatepoint_i$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_{\candidatepoint_i} \cap E_{\candidatepoint_j} = \{\}$ for all $\candidatepoint_j \in \pointbatch_k$.
	\item We make sure that $\spatialvol(\candidatepoint_i)$ is not too small, by checking whether $\spatialvol(\candidatepoint_i) > \epsilon_V \spatialvol_\text{max}$. Here $\spatialvol_\text{max}$ is the largest value of $\spatialvol(\candidatepoint_j)$ over all points $q$ in the initial set of candidate points, and $\epsilon_V$ is a small threshold parameter (we use $\epsilon_V=10^{-5}$).
	\item We make sure that all eigenvalues of $\spatialcov(\candidatepoint_i)$ are positive, and the aspect ratio of $E_{\candidatepoint_i}$ (square root of the ratio of the largest eigenvalue of $\spatialcov(\candidatepoint_i)$ to the smallest) is bounded by a constant $1/\epsilon_\Sigma$ (we use $1/\epsilon_\Sigma=20$). Negative integral kernel entries due to discretization error can cause $\spatialcov(\candidatepoint_i)$ to be indefinite or highly ill-conditioned.
\end{enumerate}
If $\candidatepoint_i$ passes these checks 
%(i.e., if $\candidatepoint_i$ is sufficiently far from other points in the batch, and $\spatialvol(\candidatepoint_i)$ and $\spatialcov(\candidatepoint_i)$ are acceptable)
then we add $\candidatepoint_i$ to $\pointbatch_k$. Otherwise we discard $\candidatepoint_i$. This process repeats until there are no more points in $\candidatepts$.  
We repeat the point selection process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots, \pointbatch_{\nbatch}$. For each batch, $\candidatepts$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem, except for points in previous batches.

@@

Here we describe how to rapidly evaluate arbitrary entries of an approximation to the integral kernel by performing radial basis function interpolation of translated and scaled versions of nearby known impulse responses. In Section~\ref{sec:h_matrix_construction_short} we use this procedure for rapidly evaluating kernel entries to construct the H-matrix approximation of $\mathbf{A}$.

Given $(y,x)\in \Omega \times \Omega$, let
$
	z_i := y - \spatialmean(x) + \spatialmean(x_i)
$
and define
\begin{equation}
\label{eq:fxyxp}
f_i := \frac{\spatialvol(x)}{\spatialvol(x_i)}\phi_{x_i}\left(z_i\right)
\end{equation}
for $i=1,\dots,\numnbr$, where $\{x_i\}_{i=1}^{\numnbr}$ are the $\numnbr$ nearest sample points to $x$, excluding points $x_i$ for which $z_i \notin \Omega$. Here $\numnbr$ is a small user-defined parameter, e.g., $\numnbr=10$. We find the $\numnbr$ nearest sample points to $x$ by querying a precomputed kd-tree~\cite{Bentley75} of all sample points.  We check whether $z_i \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree)~\cite{Ericson04} of the mesh cells used to discretize the problem. 
Note that $\phi_{x_i}\left(z_i\right)$ is well-defined because $z_i \in \Omega$, and $\frac{\spatialvol(x)}{\spatialvol(x_i)}$ is well-defined because the sample point choosing procedure in Section~\ref{sec:sample_point_selection} ensures that $\spatialvol(x_i)>0$. Per the discussion in Section~\ref{sec:local_mean_displacement_invariance}, we expect $\Aker(y,x) \approx f_i$ for $i=1,\dots,\numnbr$. The closer $x_i$ is to $x$, the better we expect the approximation to be. We therefore approximate $\Aker(y,x)$ by interpolating the (point,value) pairs
$
\left\{\left(x_i, f_i\right)\right\}_{i=1}^{\numnbr}
$
at the point $x$. 
Interpolation is performed using the following radial basis function~\cite{Wendland04} scheme:
\begin{equation}
\Aker(y,x) \approx \widetilde{\Aker}(y,x) := \sum_{i=1}^{\numnbr} \rbfweight_i~ \rbf\left(\|x-x_i\|\right),
\end{equation}
where $\rbfweight_i$ are weights, and 
$\rbf(r):= \exp\left(-\frac{1}{2}\left(\epsilon\frac{r}{L}\right)^2\right)$
is a Gaussian kernel radial basis function. Here $L:=\diam\left(\{x_i\}_{i=1}^{\numnbr}\right)$ is the diameter of the set of sample points used in the interpolation, and $\epsilon$ is a user-defined shape parameter that controls the width of the kernel function. The vector of weights, $\rbfweight = (\rbfweight_1, \rbfweight_2, \dots, \rbfweight_{\numnbr})^T$, is found as the solution to the $\numnbr \times \numnbr$ linear system
\begin{equation}
B \rbfweight = f,
\end{equation}
where $B \in \mathbb{R}^{\numnbr \times \numnbr}$,
$
B_{ij} := \rbf\left(\|x_i - x_j\|\right),
$
and $f \in \mathbb{R}^{\numnbr}$ has entries $f_i$ from~\eqref{eq:fxyxp}.

@@

We form an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$ by forming an H-matrix representation $\mathbf{\Aker}_H$ of $\mathbf{\Aker}$ then multiplying $\mathbf{\Aker}$ with mass matrices $\massmatrix$ per~\eqref{eq:boldA} to form
$
\mathbf{A}_H = \massmatrix \mathbf{\Aker}_H \massmatrix.
$
Here we use a diagonal lumped mass matrix, so these matrix-matrix multiplications are trivial. If a non-diagonal mass matrix is used, one may form a H-matrix representation of the mass matrix, then perform the matrix-matrix multiplications in \eqref{eq:boldA} using H-matrix methods.
We use H1 matrices in the numerical results, but any other H-matrix format could be used instead. For more details on H-matrices, see~\cite{Hackbusch15}.

We form $\mathbf{\Aker}_H$ using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package~\cite{Kriemann08}. For details about the algorithms used for geometrical clustering, H-matrix construction, and H-matrix operations in HLIBpro, we refer the reader to~\cite{BormGrasedyckHackbusch03,GrasedyckKriemannLeBorne08,Kriemann13}. Although $\mathbf{\Aker}$ is a dense $\fedim \times \fedim$ matrix, constructing $\mathbf{\Aker}_H$ only requires evaluation of $O(\hrank \fedim \log \fedim)$ kernel entries $\mathbf{\Aker}_{ij} = \widetilde{\Aker}(\fepoint_i, \fepoint_j)$ (see \cite{BebendorfRjasanow03}), and these entries are computed via the radial basis function interpolation method described in Section~\ref{sec:approximate_kernel_entries}. Here $\hrank$ is the rank of the highest rank block in the H-matrix. 
We emphasize that the dense matrix $\mathbf{\Aker}$ is never formed.

@@

We solve the ice sheet inverse problem with an inexact Newton preconditioned
conjugate gradient (PCG) scheme and a globalizing Armijo line search~\cite{NocedalWright99}. The Newton search directions,
$\mathbf{\searchdir}$, are obtained by solving 
%
\begin{equation}
	\label{eq:newton_system}
	\mathbf{H} \mathbf{\searchdir} = - \mathbf{g} \qquad \text{or} \qquad \mathbf{H}_\text{gn} \mathbf{\searchdir} = - \mathbf{g},
\end{equation}
%
wherein we choose the initial guess as the discretization of the constant
function $q_0$. Here $\mathbf{g}$, $\mathbf{H}$ and $\mathbf{H}_\text{gn}$ are
the discretized gradient, Hessian, and Gauss-Newton Hessian of the
inverse problem objective function, respectively, evaluated at the current Newton iterate.
To ensure positive definiteness of the Hessian we use
$\mathbf{H}_\text{gn}$ for the first five iterations, and $\mathbf{H}$ for all
subsequent iterations. The Newton iterations are terminated when
$\|\mathbf{g}\| < 10^{-6} \|\mathbf{g}_0\|$, where $\mathbf{g}_0$ is the gradient evaluated at the initial guess. Systems~\eqref{eq:newton_system} are solved inexactly using an inner PCG
iteration, which is terminated early based on the Eisenstat-Walker \cite{EisenstatWalker96} and Steihaug \cite{Steihaug83} conditions. The inverse problem governed by the advection-diffusion PDE is linear, hence Newton's method converges in one iteration. In this case the Newton linear system,~\eqref{eq:newton_system}, is solved using PCG, using termination tolerances described in Section \ref{sec:adv}.

>>>>

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here, $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. By ``non-negative integral kernel,'' we mean that entries of $\Aop$'s integral kernel 
%(defined in Equation \ref{eq:kernel_representation})
are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$. Such operators appear, for instance, as Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{BorziSchulz11,Delosreyes15,HinzePinnauUlbrichUlbrich08}, 
%AntilEtAl18
Schur complements in Schur complement methods for solving partial differential equations and Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps)~\cite{ChanMathew94,SmithBjorstadGropp96,ToselliWidlund04}, covariance operators in spatial statistics~\cite{ChenStein21,GentonKeyesTurkiyyah18,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}. Here, ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require solving a subproblem that involves PDEs, or performing other costly computations. 
%Matrix-free does not mean that matrices are never used; our approximation is ultimately converted into a hierarchical matrix (H-matrix), a compressed matrix format that permits fast and scalable linear algebra operations.

\begin{figure}
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[scale=0.3]{impulse_batch1.png}
		\caption{One impulse response batch}
		\label{fig:batches_intro}
	\end{subfigure}
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[scale=0.7]{mean_displacement_invariance.pdf}
		\caption{Local mean displacement invariance}
		\label{fig:mean_displacement_invariance}
	\end{subfigure}
%	\begin{subfigure}{0.32\textwidth}
%		\centering
%		\includegraphics[scale=0.34]{impulse_batch3.png}
%	\end{subfigure}
	\caption{(\ref{fig:batches_intro}) One batch, $\eta_b$, of normalized impulse responses, $\impulseresponse_{x}$, that arise from applying $\Aop$ to a weighted sum of scattered point sources (see Section \ref{sec:get_impulse_response}). Here, $\Aop$ is the ice sheet inverse problem data misfit Gauss-Newton Hessian described in Section~\ref{sec:numerical_results}. 
	Black stars are point source locations. Shading shows the magnitude of the normalized impulse responses (darker means larger function values). Dashed gray ellipses are estimated impulse response support ellipsoids based on the moment method in Section \ref{sec:intromoments}. The large circle is $\partial \Omega$. (\ref{fig:mean_displacement_invariance}) Illustration of impulse responses, $\impulseresponse_x$ and $\impulseresponse_{x'}$, corresponding to points $x$ and $x'$. The operator $\Aop$ is locally mean displacement invariant (Section \ref{sec:local_mean_displacement_invariance}) if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here, $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.
	}
	\label{fig:batches_and_mdi}
\end{figure}

@@

Let $\febasis_1, \febasis_2, \dots, \febasis_\fedim$ be a set of continuous finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, let $V_h := \Span\left(\febasis_1, \febasis_2, \dots, \febasis_\fedim\right)$ be the corresponding finite element space under the $L^2$ inner product, and let $\fepoint_i \in \mathbb{R}^\gdim$, $i=1,\dots, \fedim$ be the Lagrange nodes associated with the functions $\febasis_i$. We assume that the finite element basis satisfies the Kronecker property, i.e., $\febasis_i(\fepoint_i)=1$ and $\febasis_i(\fepoint_j)=0$ if $i\neq j$. 
For $u_h \in V_h$ we write $\mathbf{u} \in \mathbb{R}^\nsamplepts_\massmatrix$ to denote the coefficient vector for $u_h$ with respect to the finite element basis, i.e.,
$
u_h(x) = \sum_{i=1}^\fedim \mathbf{u}_i \febasis_i(x).
$
Linear functionals $\rho_h \in V_h'$ have coefficient dual vectors $\boldsymbol{\rho}\in \mathbb{R}^\nsamplepts_{\massmatrix^{-1}}$, with entries $\boldsymbol{\rho}_i = \rho_h(\febasis_i)$ for $i=1,\dots,\nsamplepts$.
Here, $\massmatrix \in \mathbb{R}^{\fedim \times \fedim}$ denotes the sparse finite element mass matrix which has entries $\massmatrix_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) \mathrm{d}x$ for $i,j=1,\dots,\fedim$.
The space $\mathbb{R}^\fedim_\massmatrix$ is $\mathbb{R}^\fedim$ with the inner product $(\mathbf{u},\mathbf{w})_\massmatrix := \mathbf{u}^T \massmatrix \mathbf{w}$, and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ is the analogous space with $\massmatrix^{-1}$ replacing $\massmatrix$. Direct calculation shows that $\mathbb{R}^\fedim_\massmatrix$ and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ are isomorphic to $V_h$ and $V_h'$ as Hilbert spaces, respectively.

@@

Let $C$, $L^i$, and $Q^{ij}$ be the following constant, linear, and quadratic functions:
\begin{equation*}
C(x) := 1, \qquad
L^i(x) := x^i, \qquad
Q^{ij}(x) := x^i x^j
\end{equation*}
for $i,j=1,\dots,\gdim$. Using the definition of $\spatialvol$ in~\eqref{eq:define_vol} and using~\eqref{eq:operator_simultaneous_ips}, we have
\begin{equation*}
	\spatialvol(x) = \int_\Omega \phi_x(y) C(y) ~ dy = \left(\phi_x, C\right)_{L^2(\Omega)} = \left(\mathcal{A}^T C\right)^*(x).
\end{equation*}
Hence, we compute $\spatialvol(x)$ for all $x$ simultaneously by applying $\Aop^T$ to $C$. Analogous manipulations show that $\spatialmean(x)$ and $\spatialcov(x)$ may be computed for all points $x$ simultaneously by applying $\Aop^T$ to the functions $L^i$ and $Q^{ij}$, respectively. We have
\begin{subequations}
	\label{eq:vol_mean_var}
	\begin{align}
	\spatialvol =& \left(\Aop^T C\right)^* \\
	\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
	\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
	\end{align}
\end{subequations}
for $i,j=1,\dots, \gdim$. Here, $u/w$ denotes pointwise division, $\left(u/w\right)(x) = u(x)/w(x)$, and $u\cdot w$ denotes pointwise multiplication, $(u \cdot w)(x) = u(x)w(x)$.

@@

We start with a finite set of candidate points $\candidatepts$ and build $\pointbatch_k$ incrementally with points selected from $\candidatepts$. For simplicity of explanation, here $\pointbatch_k$ and $\candidatepts$ are mutable sets that we add points to and remove points from. First we initialize $\pointbatch_k$ as an empty set. Then we select the candidate point $\candidatepoint_i \in \candidatepts$ that is the farthest away from all points in previous sample point batches $S_1 \cup S_2 \cup \dots \cup S_{k-1}$. Candidate points for the first batch $S_1$ are chosen randomly from $\candidatepts$.
Once $\candidatepoint_i$ is selected, we remove $\candidatepoint_i$ from $\candidatepts$. Then we perform the following checks:
\begin{enumerate}
	\item We check whether $\candidatepoint_i$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_{\candidatepoint_i} \cap E_{\candidatepoint_j} = \{\}$ for all $\candidatepoint_j \in \pointbatch_k$.
	\item We make sure that $\spatialvol(\candidatepoint_i)$ is not too small, by checking whether $\spatialvol(\candidatepoint_i) > \epsilon_V \spatialvol_\text{max}$. Here, $\spatialvol_\text{max}$ is the largest value of $\spatialvol(\candidatepoint_j)$ over all points $q$ in the initial set of candidate points, and $\epsilon_V$ is a small threshold parameter (we use $\epsilon_V=10^{-5}$).
	\item We make sure that all eigenvalues of $\spatialcov(\candidatepoint_i)$ are positive, and the aspect ratio of $E_{\candidatepoint_i}$ (square root of the ratio of the largest eigenvalue of $\spatialcov(\candidatepoint_i)$ to the smallest) is bounded by a constant $1/\epsilon_\Sigma$ (we use $1/\epsilon_\Sigma=20$). Negative integral kernel entries due to discretization error can cause $\spatialcov(\candidatepoint_i)$ to be indefinite or highly ill-conditioned.
\end{enumerate}
If $\candidatepoint_i$ passes these checks 
%(i.e., if $\candidatepoint_i$ is sufficiently far from other points in the batch, and $\spatialvol(\candidatepoint_i)$ and $\spatialcov(\candidatepoint_i)$ are acceptable)
then we add $\candidatepoint_i$ to $\pointbatch_k$. Otherwise we discard $\candidatepoint_i$. This process repeats until there are no more points in $\candidatepts$.  
We repeat the point selection process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots, \pointbatch_{\nbatch}$. For each batch, $\candidatepts$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem, except for points in previous batches.

@@

Here, we describe how to rapidly evaluate arbitrary entries of an approximation to the integral kernel by performing radial basis function interpolation of translated and scaled versions of nearby known impulse responses. In Section~\ref{sec:h_matrix_construction_short} we use this procedure for rapidly evaluating kernel entries to construct the H-matrix approximation of $\mathbf{A}$.

Given $(y,x)\in \Omega \times \Omega$, let
$
	z_i := y - \spatialmean(x) + \spatialmean(x_i)
$
and define
\begin{equation}
\label{eq:fxyxp}
f_i := \frac{\spatialvol(x)}{\spatialvol(x_i)}\phi_{x_i}\left(z_i\right)
\end{equation}
for $i=1,\dots,\numnbr$, where $\{x_i\}_{i=1}^{\numnbr}$ are the $\numnbr$ nearest sample points to $x$, excluding points $x_i$ for which $z_i \notin \Omega$. Here, $\numnbr$ is a small user-defined parameter, e.g., $\numnbr=10$. We find the $\numnbr$ nearest sample points to $x$ by querying a precomputed kd-tree~\cite{Bentley75} of all sample points.  We check whether $z_i \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree)~\cite{Ericson04} of the mesh cells used to discretize the problem. 
Note that $\phi_{x_i}\left(z_i\right)$ is well-defined because $z_i \in \Omega$, and $\frac{\spatialvol(x)}{\spatialvol(x_i)}$ is well-defined because the sample point choosing procedure in Section~\ref{sec:sample_point_selection} ensures that $\spatialvol(x_i)>0$. Per the discussion in Section~\ref{sec:local_mean_displacement_invariance}, we expect $\Aker(y,x) \approx f_i$ for $i=1,\dots,\numnbr$. The closer $x_i$ is to $x$, the better we expect the approximation to be. We therefore approximate $\Aker(y,x)$ by interpolating the (point,value) pairs
$
\left\{\left(x_i, f_i\right)\right\}_{i=1}^{\numnbr}
$
at the point $x$. 
Interpolation is performed using the following radial basis function~\cite{Wendland04} scheme:
\begin{equation}
\Aker(y,x) \approx \widetilde{\Aker}(y,x) := \sum_{i=1}^{\numnbr} \rbfweight_i~ \rbf\left(\|x-x_i\|\right),
\end{equation}
where $\rbfweight_i$ are weights, and 
$\rbf(r):= \exp\left(-\frac{1}{2}\left(\epsilon\frac{r}{L}\right)^2\right)$
is a Gaussian kernel radial basis function. Here, $L:=\diam\left(\{x_i\}_{i=1}^{\numnbr}\right)$ is the diameter of the set of sample points used in the interpolation, and $\epsilon$ is a user-defined shape parameter that controls the width of the kernel function. The vector of weights, $\rbfweight = (\rbfweight_1, \rbfweight_2, \dots, \rbfweight_{\numnbr})^T$, is found as the solution to the $\numnbr \times \numnbr$ linear system
\begin{equation}
B \rbfweight = f,
\end{equation}
where $B \in \mathbb{R}^{\numnbr \times \numnbr}$,
$
B_{ij} := \rbf\left(\|x_i - x_j\|\right),
$
and $f \in \mathbb{R}^{\numnbr}$ has entries $f_i$ from~\eqref{eq:fxyxp}.

@@

We form an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$ by forming an H-matrix representation $\mathbf{\Aker}_H$ of $\mathbf{\Aker}$ then multiplying $\mathbf{\Aker}$ with mass matrices $\massmatrix$ per~\eqref{eq:boldA} to form
$
\mathbf{A}_H = \massmatrix \mathbf{\Aker}_H \massmatrix.
$
Here, we use a diagonal lumped mass matrix, so these matrix-matrix multiplications are trivial. If a non-diagonal mass matrix is used, one may form a H-matrix representation of the mass matrix, then perform the matrix-matrix multiplications in \eqref{eq:boldA} using H-matrix methods.
We use H1 matrices in the numerical results, but any other H-matrix format could be used instead. For more details on H-matrices, see~\cite{Hackbusch15}.

We form $\mathbf{\Aker}_H$ using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package~\cite{Kriemann08}. For details about the algorithms used for geometrical clustering, H-matrix construction, and H-matrix operations in HLIBpro, we refer the reader to~\cite{BormGrasedyckHackbusch03,GrasedyckKriemannLeBorne08,Kriemann13}. Although $\mathbf{\Aker}$ is a dense $\fedim \times \fedim$ matrix, constructing $\mathbf{\Aker}_H$ only requires evaluation of $O(\hrank \fedim \log \fedim)$ kernel entries $\mathbf{\Aker}_{ij} = \widetilde{\Aker}(\fepoint_i, \fepoint_j)$ (see \cite{BebendorfRjasanow03}), and these entries are computed via the radial basis function interpolation method described in Section~\ref{sec:approximate_kernel_entries}. Here, $\hrank$ is the rank of the highest rank block in the H-matrix. 
We emphasize that the dense matrix $\mathbf{\Aker}$ is never formed.

@@

We solve the ice sheet inverse problem with an inexact Newton preconditioned
conjugate gradient (PCG) scheme and a globalizing Armijo line search~\cite{NocedalWright99}. The Newton search directions,
$\mathbf{\searchdir}$, are obtained by solving 
%
\begin{equation}
	\label{eq:newton_system}
	\mathbf{H} \mathbf{\searchdir} = - \mathbf{g} \qquad \text{or} \qquad \mathbf{H}_\text{gn} \mathbf{\searchdir} = - \mathbf{g},
\end{equation}
%
wherein we choose the initial guess as the discretization of the constant
function $q_0$. Here, $\mathbf{g}$, $\mathbf{H}$ and $\mathbf{H}_\text{gn}$ are
the discretized gradient, Hessian, and Gauss-Newton Hessian of the
inverse problem objective function, respectively, evaluated at the current Newton iterate.
To ensure positive definiteness of the Hessian we use
$\mathbf{H}_\text{gn}$ for the first five iterations, and $\mathbf{H}$ for all
subsequent iterations. The Newton iterations are terminated when
$\|\mathbf{g}\| < 10^{-6} \|\mathbf{g}_0\|$, where $\mathbf{g}_0$ is the gradient evaluated at the initial guess. Systems~\eqref{eq:newton_system} are solved inexactly using an inner PCG
iteration, which is terminated early based on the Eisenstat-Walker \cite{EisenstatWalker96} and Steihaug \cite{Steihaug83} conditions. The inverse problem governed by the advection-diffusion PDE is linear, hence Newton's method converges in one iteration. In this case the Newton linear system,~\eqref{eq:newton_system}, is solved using PCG, using termination tolerances described in Section \ref{sec:adv}.
